---
title: "Project 2"
author: "Name: Kira Degelsmith  \n Partner: Felix Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---
```{r setup, include = FALSE}

#### Load necessary packages ####
packages <- c("knitr", "readr", "dplyr", "leaflet", "cowplot", "tidyr", "ggplot2", "tidyverse")

installed_packages <- rownames(installed.packages())
for (p in packages) {
  if (!(p %in% installed_packages)) {
    install.packages(p)
  }
}

# Load all packages
for (p in packages) {
  library(p, character.only = TRUE)
}

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
library(readr)
library(dplyr)
library(leaflet)
library(cowplot)
library(tidyverse)
```

# Data

## WHO Global Data
Added by Kira Degelsmith
```{bash fetch, include = FALSE}

curl -L -o confirmed_global.csv https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv

curl -L -o deaths_global.csv https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv

curl -L -o recovered_global.csv https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv

```

## USPRT US Data  
Added by Felix Lopez  
```{bash fetch_us, include = FALSE}
curl -L -o time_series_covid19_confirmed_US.csv https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv

curl -L -o time_series_covid19_deaths_US.csv https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv

```

## Read Data
```{r data}
confirmed_global <- read_csv('confirmed_global.csv')
deaths_global <- read_csv('deaths_global.csv')
recovered_global <- read_csv('recovered_global.csv')

confirmed_us <- read_csv("time_series_covid19_confirmed_US.csv")
deaths_us <- read_csv("time_series_covid19_deaths_US.csv")

## (Felix)

# Get latest date column
latest_date_us <- names(confirmed_us)[ncol(confirmed_us)]

# Group US confirmed by state
confirmed_us_summary <- confirmed_us %>%
  group_by(Province_State) %>%
  summarise(total_confirmed = sum(!!sym(latest_date_us), na.rm = TRUE))

# Group US deaths by state
deaths_us_summary <- deaths_us %>%
  group_by(Province_State) %>%
  summarise(total_deaths = sum(!!sym(latest_date_us), na.rm = TRUE))

# Merge
us_summary <- left_join(confirmed_us_summary, deaths_us_summary, by = "Province_State")

# Preview of table
head(us_summary)

##The code above downloads and processes the US COVID-19 confirmed and death data. It extracts the most recent totals for each state and merges the two datasets into a unified summary. 

```


# Project Objectives

## Objective 1
```{r ob1}

```

## Objective 2 - 

```{r ob2}

```

## Objective 3 Zooming Into Our State (Felix)
We filtered the dataset to focus on California and reshaped the data to show daily confirmed COVID-19 cases. A statewide trend plot was created, marking key milestones: second vaccine dose, Delta, and Omicron variants. We then identified the top three counties by case count and plotted their trends over time. The final visualization compares statewide and county-level patterns side by side.

```{r ob3}

# Filter for California
ca_data <- confirmed_us %>%
  filter(Province_State == "California")

# Convert the data to long format
ca_long <- ca_data %>%
  pivot_longer(cols = matches("^\\d+/\\d+/\\d+$"), 
               names_to = "Date", 
               values_to = "Confirmed") %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))
# Sum confirmations across all counties for each date
ca_statewide <- ca_long %>%
  group_by(Date) %>%
  summarize(Total_Confirmed = sum(Confirmed))

# Key dates
second_dose <- as.Date("2021-03-12")
delta_date <- as.Date("2021-05-11")
omicron_date <- as.Date("2021-11-26")

# Create statewide plot
p1 <- ggplot(ca_statewide, aes(x = Date, y = Total_Confirmed)) +
  geom_point(color = "blue", size = 0.5) +
  geom_vline(xintercept = as.numeric(second_dose), color = "purple", linetype = "dashed") +
  geom_vline(xintercept = as.numeric(delta_date), color = "orange", linetype = "dashed") +
  geom_vline(xintercept = as.numeric(omicron_date), color = "red", linetype = "dashed") +
  labs(title = "California COVID-19 Confirmations Over Time",
       x = "Date", y = "Total Confirmations")

# Find top 3 counties by max confirmation
top3_counties <- ca_long %>%
  group_by(Admin2) %>%
  summarize(Max_Confirmed = max(Confirmed)) %>%
  arrange(desc(Max_Confirmed)) %>%
  slice(1:3) %>%
  pull(Admin2)

# Filter for top 3 counties
top3_data <- ca_long %>%
  filter(Admin2 %in% top3_counties)

# Plot top 3 counties
p2 <- ggplot(top3_data, aes(x = Date, y = Confirmed, color = Admin2)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = as.numeric(second_dose), color = "purple", linetype = "dashed") +
  geom_vline(xintercept = as.numeric(delta_date), color = "orange", linetype = "dashed") +
  geom_vline(xintercept = as.numeric(omicron_date), color = "red", linetype = "dashed") +
  labs(title = "Top 3 CA Counties COVID-19 Confirmations",
       x = "Date", y = "Confirmed Cases") +
  theme(legend.position = "bottom", legend.title = element_blank())

plot_grid(p1, p2, labels = c("A", "B"))

```

## Objective 4

```{r ob4}
# Merge population data from deaths_us (which contains Population column)
us_summary <- us_summary %>%
  left_join(deaths_us %>% select(Province_State, Population), by = "Province_State") %>%
  filter(Population > 0, total_confirmed > 0, total_deaths > 0)

# Add log-transformed columns
us_summary <- us_summary %>%
  mutate(
    log_population = log10(Population),
    log_confirmed = log10(total_confirmed),
    log_deaths = log10(total_deaths)
  )

# Scatter plot: Population vs Confirmed Cases
pop_vs_confirmed <- ggplot(us_summary, aes(x = log_population, y = log_confirmed)) +
  geom_point(color = "darkblue") +
  labs(
    title = "Log(Population) vs Log(Confirmed Cases)",
    x = "Log10 Population",
    y = "Log10 Confirmed Cases"
  ) +
  theme_minimal()

# Scatter plot: Confirmed Cases vs Deaths
confirmed_vs_deaths <- ggplot(us_summary, aes(x = log_confirmed, y = log_deaths)) +
  geom_point(color = "darkred") +
  labs(
    title = "Log(Confirmed Cases) vs Log(Deaths)",
    x = "Log10 Confirmed Cases",
    y = "Log10 Deaths"
  ) +
  theme_minimal()

# Combine plots using cowplot
plot_grid(pop_vs_confirmed, confirmed_vs_deaths, labels = c("A", "B"))

```

### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```





